HMM

### 2016-01-13 paper3.Rnw
updates

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in behavioural mode transition 
\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in transitioning behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\begin{table}[]
\centering
\caption{My caption}
\label{Cat-ID}
\begin{tabular}{ccc}
 (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
  130 & 1 & 10286    \\
  131& 2  & 1111    \\
 48& 14  & 12222    \\
 94& 15  & 9999  
\end{tabular}
\end{table}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever) ML: What do you think? Are ecologist more comfortable with modes/ latent states?}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which movement modes
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

@@

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting. \todo{where do we actually focus on these three states?  Think this could probably be introduced in the Discussion, if we need it}

>>>>

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in transitioning behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\begin{table}[]
\centering
\caption{Cat-ID}
\label{Cat-ID}
\begin{tabular}{ccc}
 (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
  130 & 1 & 10286    \\
  131 & 2  & 9458    \\
 48& 14  & 14645    \\
 94& 15  & 10250  
\end{tabular}
\end{table}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever) ML: latent states}
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which latent states
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

@@

We considered 3 (biologically interpretable number of states identified: (1) resting state, characterized by short step length; (2) Moderately active state characterized by intermediate step lengths, and (3) Traveling state, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting. \todo{where do we actually focus on these three states?  Think this could probably be introduced in the Discussion, if we need it}
